---
title: "Machine Learning Project"
author: "Darcy Lewis"
date: "June 5, 2015"
output: html_document
---

# Overview
Objective of this analysis is to build a model which can predict the manner (A, B, C, D or E) in which the individuals performed bar bell lifts, given data recorded by accelerometers attached to the individuals.

The variable "classe"" in the data set correctly documents the manner in which the lift was performed.

Model Build Choices: 

1. Pre-Processing & Predictor Elimination
With 159 possible predictors, wanted to reduce the number of predictors to those most likely prior to attempting to find a model to predict the outcome.  The steps to identifing good predictors took the field from 159 down to 14.  
 - used only numeric fields; 56 predictors remaining  
 - removed those columns which were contextual such as datetime stamps and row numbers; 52 predictors remaining  
 - removed near zero columns (after replacing NAs with 0, replacing missing values with zero and centering and scaling the data); 26 predictors remaining
 - removed highly correlated predictors (>75%); 24 predictors remaining
 - looked for predictors which were linear combinations of oneanother, but none were found  
Note: centering and scaling was necessary as data valuess in various predictors ranged from fractions to values in the low hundreds  

2. Selecting a Model
 - linear regression was not an option as the outcome variable had 5 possible values  
 - decision trees allow for more than two possible outcomes, but 40% accuracy just on the training data    
 - random forest resulted in 100% accuracy on the training data and <5% out of sample error rate. 
  
##1 Setup Environment
```{r}
library(caret)
library(ggplot2)
library(rpart)
```
##2 Import Training Data
```{r}
setwd("C:/Users/Darcy/Documents/Coursera/Machine Learning")
#fileUrl<-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
#download.file(fileUrl,destfile=".",method="curl")
pml_data<-read.csv("pml-training.csv")
```
##3 Prepare for Cross Validation
```{r}
##Partition the training data into training and test (80/20 ratio) data sets 
set.seed(1234)
inTrain<-createDataPartition(y=pml_data$classe,p=0.8,list=FALSE)
training<-pml_data[inTrain,]
testing<-pml_data[-inTrain,]
```
##4 Pre-Processing & Predictor Elimination
```{r}
##create data frame of just the potential numeric predictors
NumericPredictors <- sapply(training, is.numeric)
trainingPredictors<-training[ , NumericPredictors]

##Remove the 1st 4 columns containing contextual data
trainingPredictors <- trainingPredictors[,5:56]    

##replace NA values with 0
trainingPredictors[is.na(trainingPredictors)]<-0

##center and scale the data
preProcValues <- preProcess(trainingPredictors, method = c("center", "scale"))
trainingNorm <- predict(preProcValues, trainingPredictors)

#remove near zero columns from the training dataset after replacing NA values with 0 to create dataframe of NonZero (NZ) predictors
nzv <- nearZeroVar(trainingNorm)
trainingNZ <- trainingNorm[, -nzv]

##remove highly correlated predictors
CorTrainingMatrix<-cor(trainingNZ)
highlyCorPredictors <- findCorrelation(CorTrainingMatrix, cutoff = 3/4)
trainingRelevant <- trainingNZ[,-highlyCorPredictors]

##append outcome variable back into final dataset to be modeled, i.e. trainingClean
classe<-training$classe
trainingClean<-cbind(trainingRelevant,classe)
```
##Create an Accurate Predictive Model
```{r}
##After trying several models, random forest gave by far the best results
modelFit<-train(classe~.,method="rf",data=trainingClean,prox=TRUE)
confusionMatrix(trainingClean$classe,predict(modelFit,trainingClean))
```
##Pre-process and Apply the Model to Test Partition
```{r}
##preprocess testing data using training data
testingPredictors<-subset(testing,select=colnames(trainingPredictors))
testingPredictors[is.na(testingPredictors)]<-0
testingNorm <- predict(preProcValues, testingPredictors)
classe<-testing$classe
testingNorm<-cbind(testingNorm,classe)
testingClean<-subset(testingNorm,select=colnames(trainingClean))
##apply the model to the testing partion
confusionMatrix(testingClean$classe,predict(modelFit,testingClean))
```